Bicycle Balance Assistance Reduces Probability of Falling at Low Speeds When Subjected to Mechanical Perturbations

### 2024-06-27 main.tex
More improvements to the discussion section.

In the early years of developments in automatic control, \citep{Whipple1899}
not only derived the correct equations of motion of the bicycle but realized
and showed that roll motion feedback can stabilize a bicycle. Much later,
attempts at automatic roll stabilization of single track vehicles began in
earnest after predictive motorcycle models were developed and refined in
throughout the 1970s. \Citet{vanZytveld1975} was influenced by Whipple's work
and seems to be the first to attempt to robotically stabilize a small motorbike
with a controlled inverted pendulum that mimicked rider lean, but he was not
successful in demonstrating what his theoretic control model correctly
predicted. In his model, he recognized that feedback of the vehicle roll angle
and angular rate was essential to stabilize the vehicle. It was not until the
early 1980s that \cite{Nagai1983} successfully demonstrated balancing a robotic
bicycle on a treadmill with both steering control and a laterally moving mass.
Ruijs and Pacjecka followed this by demonstrating an automatically balanced
motorcycle~\citep{Ruijs1986} and they did so with a steering motor. Ruijs and
Pacejka clearly showed that steer torque driven by roll angle feedback
stabilizes the capsize mode, roll angular rate feedback stabilizes the weave
mode, and steer angular rate feedback stabilizes the wobble mode. They also
proposed how the gains must change with respect to vehicle speed for favorable
control at all speeds. This roll motion feedback is the simplest controller
that can stabilize a single track vehicle above a minimum speed when one is not
concerned with wobble instabilities, something not prominent in bicycles at low
speeds. Ruijs and Pacejka's work was not particularly concerned with low speed
stability and their vehicle was fully automatic, no human rider was involved.
Many more automatically balanced single track vehicles have been demonstrated
over the last 40 years, but none have demonstrated that increasing low speed
stability can assist in balancing and possibly reduce single actor falls. Most
of these robotic bicycle and motorcycle designers did not intend for a human
rider to also control the stabilized vehicle. Nevertheless, an automatically
stabilized bicycle can be controlled by a human rider if the motor controlled
steer torque and the rider applied steer torque act on the steer in parallel.
The effect of this automatic control gives the ability to change the dynamics,
up to some limits, of the human controlled plant. In our prior
study~\citep{Alizadehsaravi2023}, we demonstrate reduced motion during
distractions due to the balance assist but contrarily researchers with a
similar vehicle recently showed rider dissatisfaction with the
stabilization~\cite{Hanakam2023}. \todo{Maybe this paragraph can be a bit 
shorter? Of course important to mention previous research, but all this 
info is not 100% necessary for understanding our paper.}

The linear Carvallo-Whipple bicycle model~\citep{Carvallo1899,Whipple1899} is
the simplest bicycle model that exhibits non-minimum phase behavior and
self-stability. The bicycle model is ideal for showing the effect of a roll
motion feedback driven steer motor on the dynamics. The linear model can be
described by the fourth order state space equations:
%
\begin{align}
  \dot{\vec{x}} = \mathbf{A} \vec{x} +
  \mathbf{B}
  \vec{u}
  \textrm{ where }
  \vec{x} = \begin{bmatrix} \phi \\ \dot{\phi} \\ \delta \\ \dot{\delta} \end{bmatrix}
  \textrm{ and }
  \vec{u} = \begin{bmatrix} T_{\phi} \\ T_{\delta} \end{bmatrix}
  \textrm{.}
\end{align}
%
The states are the roll angle \(\phi\) and steer angle \(\delta\) along with
their derivatives and the inputs are roll torque \(T_\phi\) and steer torque
\(T_\delta\). The state \(\mathbf{A}\) and input \(\mathbf{B}\) matrices are
functions of the equilibrium forward speed \(v\) and  are otherwise populated
with expressions that are functions of the geometric and inertial parameters of
the nonholonomic multibody system made up of four rigid bodies: two wheels,
front frame, and rear frame.

@@

\subsection{Perturbations}
%
We apply external longitudinal forces to the ends of each handlebar using an
adapted Bump'Em system~\cite{Tan2020} which is designed with four motors
working in tandem, Figure~\ref{fig:setup-diagram}. The four motors are
programmed to apply a light force at all times to keep the ropes taught and to
track a commanded force profile using a PID controller running on a
microprocessor (Arduino Mega 2560, Italy). We measure the force applied by each
motor at the handlebar via four inline load cells, each with a max load of
250~\si{\newton}. The left rear load cell  was damaged during pilot experiments, and is
replaced by a load cell with a max load of 500~\si{\newton}. The commanded
force profiles are designed to apply an external pulsive torque to the front
assembly (handlebars, forks, wheel) at magnitudes varying from 20 to 200~\si{\Newton} per motor. 
This results in a torque between 16 and 160~\si{\Newton\meter}. The
four motors are arranged at the four corners of a 1~\si{\meter} wide treadmill
(EC-90 Flat, Maxon Group, Switzerland) that can reach speeds of 18~\si{\kilo\meter\per\hour}.
The general design is described in detail in the Van De Velde's MSc
hesis~\citep{vandeVelde2022a} and the physical arrangement  is shown in
Figure~\ref{fig:participant-in-set-up}. Our modifications relative to that
described in the thesis include simplifying the controller with a inexpensive
microcontroller and the use of a simpler safety harness.
Figure~\ref{fig:perturbation_0} shows an example perturbation.
%
% NOTE : I get a compile error with the \protect calls here, not sure why.
\begin{figure}
  \centering
  \subcaptionbox{A participant on the bicycle in the safety harness with the
    Bump'Em motor lines attached to the ends of the
    handlebars.\protect\label{fig:participant-in-set-up}}
    {\includegraphics[width=75mm]{figures/participant-in-set-up.png}}
  \subcaptionbox{Top view diagram of bicycle handlebar perturbation
    system.\protect\label{fig:setup-diagram}}
    {\includegraphics[width=65mm]{figures/setup-diagram.png}}
  \caption{Diagrams showing the Bump'Em system arranged to apply handlebar
    perturbations.}
  \label{fig:experimental-setup}
\end{figure}

@@

The magnitude of the perturbation is determined using an adaptive staircase procedure. 
\todo{Cite to correct paper from thesis}
The adaptive staircase takes the nominal fall threshold as input and randomly chooses 
one of five possible forces that are centred around the nominal fall threshold with 
10~\si{\Newton} intervals. If the perturbation results in a fall, the nominal fall 
threshold is decreased with 10~\si{\Newton}. If the perturbation does not results in a 
fall, the nominal fall threshold is increased with 10~\si{\Newton}. This adaptive
staircase procedure ensures that approximately 50\% of the perturbations applied to the
rider result in a fall. 
%
\begin{figure}
  \centering
  \subcaptionbox{Start}{\includegraphics[width=40mm]{figures/start.png}}
  \subcaptionbox{During}{\includegraphics[width=40mm]{figures/during.png}}
  \subcaptionbox{After}{\includegraphics[width=40mm]{figures/after.png}}
  \subcaptionbox{Recovery}{\includegraphics[width=40mm]{figures/recovery.png}}
  \caption{Video frames depicting the application of a perturbation and the
  rider's response and recovery.}
  \label{fig:perturbation-sequence}
\end{figure}

@@

\begin{figure}
  \centering
  \subcaptionbox{1.7~\si{\mps}\label{fig:probability-6kph}}{
    \includegraphics[width=120mm]{figures/predicted_fall_probability_6kmh.png}
  }
  \subcaptionbox{2.8~\si{\mps}}{
    \includegraphics[width=120mm]{figures/predicted_fall_probability_10kmh.png}
  }
  \caption{Comparison of predicted fall probability for balance-assist on or
    off All other predictor variables are fixed at zero, except for the
    interaction effect between the balance-assist and the angular impulse. The
    x axis is standard deviation around the mean (all perturbations), centered
    per participant.}
  \label{fig:probability}
\end{figure}

\section{Discussion}
%
Starting with Figure~\ref{fig:probability} and Table~\ref{tab:freq-coeffs-6}.

To interpret the results in Tables~\ref{tab:freq-coeffs-6} and
\ref{tab:freq-coeffs-10}
it is important to understand the relationship between probability and odds.
The probability that a fall occurs depends on the values of all the indepedent
variables and but we can visualize one view into the results in
Figure~\ref{fig:probability}. The estimate in Table~\ref{tab-freq-coeffs-6}
shows that if the odds that the balance assist system halves the odds that a
perturbation results in a fall. For example, if the odds are a 1000:1,
turning on the balance-assist system reduces the odds to 500:1. However, the
probability that
a fall occurs is only reduced from $0.999$ to $0.998$. If
the odds that a fall will occur are smaller, halving the odds has a larger
influence on the fall probability.For example, if the odds that a fall occurs
is two, halving it to one reduces the probability from $0.66$ to $0.50$. This
clearly shown in the behavior in Figure~\ref{fig:probablity} with the behavior
at high standard deviations.

@@

Angular impulse magnitude has the largest significant effect for predicting
fall probability. An increase in angular impulse increases the fall probability
both at 6 and 10~\si{\kph}. At 10~\si{\kph}, the multiplicative change in odds
is approximately twice as big as at 6~\si{\kph}. Thus, angular impulse is a
more important predictor at higher speeds compared to lower speeds. The reason
for this likely has to do with the width of the treadmill. As a bicycle travels
at higher speeds, the same perturbation causes larger lateral deviations. At
10~\si{\kph} almost all falls were due to the bicycle exiting the maximum width
of the treadmill. If the same experiment was performed on an infinite plane,
the riders may have recovered from the perturbation. At 6~\si{\kph} the riders
could often recover in the allotted treadmill width. Our results are very much
dependent on the two modes of falling with use: exit the treadmill width or
foot is placed on the belt. Cycle paths are a similar width as the treadmill,
so ride's are often limited in width when recovering from a fall.

@@

\section{Conclusion}

>>>>

In the early years of developments in automatic control, \citep{Whipple1899}
not only derived the correct equations of motion of the bicycle but realized
and showed that roll motion feedback can stabilize a bicycle. Much later,
attempts at automatic roll stabilization of single track vehicles began in
earnest after predictive motorcycle models were developed and refined in
throughout the 1970s. \Citet{vanZytveld1975} was influenced by Whipple's work
and seems to be the first to attempt to robotically stabilize a small motorbike
with a controlled inverted pendulum that mimicked rider lean, but he was not
successful in demonstrating what his theoretic control model correctly
predicted. In his model, he recognized that feedback of the vehicle roll angle
and angular rate was essential to stabilize the vehicle. It was not until the
early 1980s that \cite{Nagai1983} successfully demonstrated balancing a robotic
bicycle on a treadmill with both steering control and a laterally moving mass.
Ruijs and Pacjecka followed this by demonstrating an automatically balanced
motorcycle~\citep{Ruijs1986} and they did so with a steering motor. Ruijs and
Pacejka clearly showed that steer torque driven by roll angle feedback
stabilizes the capsize mode, roll angular rate feedback stabilizes the weave
mode, and steer angular rate feedback stabilizes the wobble mode. They also
proposed how the gains must change with respect to vehicle speed for favorable
control at all speeds. This roll motion feedback is the simplest controller
that can stabilize a single track vehicle above a minimum speed when one is not
concerned with wobble instabilities. Ruijs and Pacejka's work was not
particularly concerned with low speed stability and their vehicle was fully
automatic, no human rider was involved. Many more automatically balanced single
track vehicles have been demonstrated over the last 40 years, but none have
demonstrated that increasing low speed stability can assist in balancing and
possibly reduce single actor falls. Most of these robotic bicycle and
motorcycle designers did not intend for a human rider to also control the
stabilized vehicle. Nevertheless, an automatically stabilized bicycle can be
controlled by a human rider if the motor controlled steer torque and the rider
applied steer torque act on the steer in parallel.  The effect of this
automatic control gives the ability to change the dynamics, up to some limits,
of the human-controlled plant. In our prior study~\citep{Alizadehsaravi2023},
we demonstrate reduced motion during distractions due to the balance assist but
contrarily researchers with a similar vehicle recently showed rider
dissatisfaction with the stabilization~\cite{Hanakam2023}. \todo{Maybe this
paragraph can be a bit shorter? Of course important to mention previous
research, but all this info is not 100\% necessary for understanding our paper.
(I've been working on this paragraph for over a year for a few different
possible publications about the balance assist bike, I'd like to keep it in.
This is an attempt to set the record straight on where this entire idea of
doing "balance assist" derives from.-- JKM)}

The linear Carvallo-Whipple bicycle model~\citep{Carvallo1899,Whipple1899} is
the simplest bicycle model that exhibits non-minimum phase behavior and
self-stability. The bicycle model is ideal for showing the effect of a roll
motion feedback driven steer motor on the dynamics. The linear version of this
model can be described by the fourth order state space
equations~\citep{Meijaard2007}:
%
\begin{align}
  \dot{\vec{x}} = \mathbf{A} \vec{x} +
  \mathbf{B}
  \vec{u}
  \textrm{ where }
  \vec{x} = \begin{bmatrix} \phi \\ \dot{\phi} \\ \delta \\ \dot{\delta} \end{bmatrix}
  \textrm{ and }
  \vec{u} = \begin{bmatrix} T_{\phi} \\ T_{\delta} \end{bmatrix}
  \textrm{.}
\end{align}
%
The states are the roll angle \(\phi\) and steer angle \(\delta\) along with
their derivatives and the inputs are roll torque \(T_\phi\) and steer torque
\(T_\delta\). The state \(\mathbf{A}\) and input \(\mathbf{B}\) matrices are
functions of the equilibrium forward speed \(v\) and  are otherwise populated
with expressions that are functions of the geometric and inertial parameters of
the nonholonomic multibody system made up of four rigid bodies: two wheels,
front frame, and rear frame.

@@

\subsection{Perturbations}
%
We apply external longitudinal forces to the ends of each handlebar using an
adapted Bump'Em system~\cite{Tan2020} which is designed with four motors
working in tandem, Figure~\ref{fig:setup-diagram}. The four motors are
programmed to apply a light force at all times to keep the ropes taught and to
track a commanded force profile using a PID controller running on a
microprocessor (Arduino Mega 2560, Italy). We measure the force applied by each
motor at the handlebar via four inline load cells, each with a max load of
250~\si{\newton}. The left rear load cell  was damaged during pilot
experiments, and is replaced by a load cell with a max load of
500~\si{\newton}. The commanded force profiles are designed to apply an
external pulsive torque to the front assembly (handlebars, forks, wheel) at
magnitudes varying from 20 to 200~\si{\newton} per motor.  This results in a
torque between 16 and 160~\si{\newton\meter}. The four motors are arranged at
the four corners of a 1~\si{\meter} wide treadmill (EC-90 Flat, Maxon Group,
Switzerland) that can reach speeds of 18~\si{\kilo\meter\per\hour}.  The
general design is described in detail in the Van De Velde's MSc
hesis~\citep{vandeVelde2022a} and the physical arrangement  is shown in
Figure~\ref{fig:participant-in-set-up}. Our modifications relative to that
described in the thesis include simplifying the controller with a inexpensive
microcontroller and the use of a simpler safety harness.
Figure~\ref{fig:perturbation_0} shows an example perturbation.
%
% NOTE : I get a compile error with the \protect calls here, not sure why.
\begin{figure}
  \centering
  \subcaptionbox{A participant on the bicycle in the safety harness with the
    Bump'Em motor lines attached to the ends of the
    handlebars.\protect\label{fig:participant-in-set-up}}
    {\includegraphics[width=75mm]{figures/participant-in-set-up.png}}
  \subcaptionbox{Top view diagram of bicycle handlebar perturbation
    system.\protect\label{fig:setup-diagram}}
    {\includegraphics[width=65mm]{figures/setup-diagram.png}}
  \caption{Diagrams showing the Bump'Em system arranged to apply handlebar
    perturbations.}
  \label{fig:experimental-setup}
\end{figure}

@@

The magnitude of the perturbation is determined using an adaptive staircase
procedure.  \todo{Cite to correct paper from thesis} The adaptive staircase
takes the nominal fall threshold as input and randomly chooses one of five
possible forces that are centred around the nominal fall threshold with
10~\si{\newton} intervals. If the perturbation results in a fall, the nominal
fall threshold is decreased with 10~\si{\newton}. If the perturbation does not
results in a fall, the nominal fall threshold is increased with
10~\si{\newton}. This adaptive staircase procedure ensures that approximately
50\% of the perturbations applied to the rider result in a fall.
%
\begin{figure}
  \centering
  \subcaptionbox{Start}{\includegraphics[width=40mm]{figures/start.png}}
  \subcaptionbox{During}{\includegraphics[width=40mm]{figures/during.png}}
  \subcaptionbox{After}{\includegraphics[width=40mm]{figures/after.png}}
  \subcaptionbox{Recovery}{\includegraphics[width=40mm]{figures/recovery.png}}
  \caption{Video frames depicting the application of a perturbation and the
  rider's response and recovery.}
  \label{fig:perturbation-sequence}
\end{figure}

@@

\begin{figure}
  \centering
  \subcaptionbox{6~\si{\kph}\label{fig:probability-6kph}}{
    \includegraphics[width=120mm]{figures/predicted_fall_probability_6kmh.png}
  }
  \subcaptionbox{2.8~\si{\mps}}{
    \includegraphics[width=120mm]{figures/predicted_fall_probability_10kmh.png}
  }
  \caption{Comparison of predicted fall probability for balance-assist on or
    off All other predictor variables are fixed at zero, except for the
    interaction effect between the balance-assist and the angular impulse. The
    x axis is standard deviation around the mean (all perturbations), centered
    per participant.}
  \label{fig:probability}
\end{figure}

\section{Discussion}
%
The probability that a fall occurs depends on the values of all the independent
variables in Table~\ref{tab:stat-model-variables}, but we can visualize the
effect of one or two variables (e.g. Figure~\ref{fig:probability}) to gain
some insight.
But to interpret the results in Tables~\ref{tab:freq-coeffs-6} and
\ref{tab:freq-coeffs-10} it is important to understand the relationship between
probability and odds.
The estimate in Table~\ref{tab-freq-coeffs-6} shows that the balance assist
system halves the odds that a perturbation results in a fall
(MCiO=0.53)~\todo{Make a variable for MCiO}.
This means if the odds are a 1000:1, turning on the balance-assist system
reduces the odds to 500:1.
However, the probability that a fall occurs is only reduced from $0.999$ to
$0.998$ in that case.
If the odds that a fall will occur are smaller, halving the odds has a larger
influence on the fall probability.
For example, if the odds that a fall occurs is two, halving it to one reduces
the probability from $0.66$ to $0.50$.
This can be seen in Figure~\ref{fig:probablity} which shows the fall
probability as a function of the magnitude of the normalized angular impulse
for both speeds.
The difference in probability between the balance-assist on/off case becomes
insignificant outside of approximately \(\pm1\) standard deviation of the
average angular impulse.
This means that the balance assist is most effective for perturbations close to
the subject's personal threshold between falling or recovery and that large
perturbations will make you fall regardless of the balance assist's help.

@@

Angular impulse magnitude has the largest significant effect for predicting
fall probability, Tables~\ref{tab:freq-coeffs-6} and \ref{tab:freq-coeffs-10}.
An increase in angular impulse increases the fall probability both at 6 and
10~\si{\kph}. At 10~\si{\kph}, the multiplicative change in odds is
approximately twice as big as at 6~\si{\kph}. Thus, angular impulse is a more
important predictor at higher speeds compared to lower speeds. The reason for
this likely has to do with the width of the treadmill and may also be why
balance assist system did not have a statistically significant effect at
10~\si{\kph}. As a bicycle travels at higher speeds, the same perturbation
causes larger lateral deviations. At 10~\si{\kph} almost all falls were due to
the bicycle exiting the maximum width of the treadmill. If the same experiment
was performed on an infinite plane, the riders may have recovered from more
perturbations. At 6~\si{\kph} the riders could often recover in the allotted
treadmill width. Our results are very much dependent on the two modes of
falling with use: exit the treadmill width or foot is placed on the belt. Cycle
paths are a similar width as the treadmill, so rider's are often limited in
width when recovering from a fall.

@@

\section{Conclusion}
%
TODO
